COVID19 - Forecast and predictions using a time dependent SEIR model

### 2020-03-15 draft_analysis_Italy_new.Rmd
Add home page, navigation and theme

```{r echo=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir= normalizePath('..'))
knitr::opts_chunk$set(error = FALSE)
```
---
title: "COVID19 - Forecast and predictions using a time dependent SEIR model"
author: "Paolo Girardi"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:html_document:
  df_print: paged
pdf_document: 
  toc: yes
---

<a rel="license" href="http://creativecommons.org/licenses/by-nc/4.0/"><img alt="Creative Commons License" style="border-width:0" src="https://i.creativecommons.org/l/by-nc/4.0/88x31.png" /></a><br />This work is licensed under a <a rel="license" href="http://creativecommons.org/licenses/by-nc/4.0/">Creative Commons Attribution-NonCommercial 4.0 International License</a>.

@@

## The COVID dataset

@@

```{r echo=TRUE}
checkpackage <- function(package) {
  if (!package %in% installed.packages()) install.packages(package)
}
checkpackage("dygraphs")
checkpackage("xts")
checkpackage("EpiDynamics")
checkpackage("webshot")
checkpackage("bsts")
```
and load them.
```{r, warning=FALSE}
library(dygraphs)
library(xts)
library(EpiDynamics)
library(webshot)
library(bsts)
```

@@

We calculate several R0 values, each one based on a mobile window of 5 days, that can be sufficient to estimate a local trend, in order to assess if the R0 trend is decreasing (how is expected to be).

@@

library(ggplot2)
fp <- ggplot(data=df, aes(x=label, y=mean, ymin=lower, ymax=upper)) +
  geom_pointrange() +
  geom_hline(yintercept=1, lty=2) +  # add a dotted line at x=1 after flip
  xlab("Date") + ylab("R0 Mean (95% CI)") +
  theme_bw() 
print(fp)

@@

The R0 shows a decreasing trend in the last period. We use the estimated trend between R0 and time to calculate the future R0 value for the next 14 days. 
We predict beta (and R0) for the next 14 days by means of a linear regressione model, assuming a Log-normal distribution for the beta (the slope) forcing its value to be greater than 0.
```{r R0 forecast}

@@

<img src="http://www.public.asu.edu/~hnesse/classes/seireqn.png"/>  
<img src="https://upload.wikimedia.org/wikipedia/commons/3/3d/SEIR.PNG"/>

We perfom simulation on the trend of beta coefficient by means of a Bayesian Structural Time Series using the library bsts of R.   
We made a forecast forward of 14 days.

```{r bsts model fit and prediction, echo=TRUE}
# Bayesian Structural Time Series
ss <- AddLocalLinearTrend(list(), log(beta_vec))
model1 <- bsts(log(beta_vec),
               state.specification = ss,
               niter = 200)
par(mfrow = c(1,1))
plot(model1, "components", ylab="Log beta coefficient", xlab="days")
#previsioni
pred1 <- predict(model1, horizon = 16, burn = 100)
par(mfrow = c(1,1))
plot(pred1 , ylab="Log beta coefficient",main="Data and predictions")
# matrix of beta coefficients
coef<-(exp(pred1$distribution)[,3:16])
par(mfrow = c(1,1))
```
For each vectors of simulated beta coefficients we perform a SEIR model (one for each scenario).  
We save the results and plot a credible interval at 90% (5-95% percentile).

```{r  scenario plot , echo=TRUE}
library(EpiDynamics)
seir1_sim<-seir2_sim<-seir3_sim<-NULL
for(s in 1:dim(coef)[1]){
  # average number of single connections of an infected person
  # less contacts, less probability of new infections
  # we keep constant the other parameters
  forecast<-14
  seir1<-seir2<-seir3<-NULL
  for(i in 1:forecast){
    parameters <- c(mu = mu0, beta = matrix(coef[s,i]), sigma = sigma0, gamma = 1/duration)
    f1<-10
    if( i==1) initials <- c(S = 0.95, E = (f1*I0/N), I = I0/N, R = R0/N)
    if( i>1) initials <- c(S = seir1_temp$results$S[2], E = seir1_temp$results$E[2], I =seir1_temp$results$I[2], R = seir1_temp$results$R[2])
    seir1_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
    seir1 <- rbind(seir1,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
    f2<-5
    if( i==1) initials <- c(S = 0.95, E = (f2*I0/N), I = I0/N, R = R0/N)
    if( i>1) initials <- c(S = seir2_temp$results$S[2], E = seir2_temp$results$E[2], I =seir2_temp$results$I[2], R = seir2_temp$results$R[2])
    seir2_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
    seir2 <- rbind(seir2,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
    f3<-3
    if( i==1) initials <- c(S = 0.95, E = (f3*I0/N), I = I0/N, R = R0/N)
    if( i>1) initials <- c(S = seir3_temp$results$S[2], E = seir3_temp$results$E[2], I =seir3_temp$results$I[2], R = seir3_temp$results$R[2])
    seir3_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
    seir3 <- rbind(seir3,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
  }
  seir1_sim<-rbind(seir1_sim,cbind(rep(s,forecast),seir1))
  seir2_sim<-rbind(seir2_sim,cbind(rep(s,forecast),seir2))
  seir3_sim<-rbind(seir3_sim,cbind(rep(s,forecast),seir3))
}
seir1_sim[,2]<-seir2_sim[,2]<-seir3_sim[,2]<-rep(1:forecast,dim(coef)[1])
colnames(seir1_sim)[1]<-colnames(seir2_sim)[1]<-colnames(seir3_sim)[1]<-"sim"

### confidence limits
I_seir_med<-tapply(seir1_sim$I,seir1_sim$time,median)
I_seir_lwr<-tapply(seir1_sim$I,seir1_sim$time,quantile,p=0.05)
I_seir_upr<-tapply(seir1_sim$I,seir1_sim$time,quantile,p=0.95)

days.before<-date[1:days]
days.ahead<-date[(days+1):(days+forecast)]
step.ahead<-forecast+1
mu.lower<-c(dat_csv$totale_attualmente_positivi,I_seir_lwr*N)
mu.upper<-c(dat_csv$totale_attualmente_positivi,I_seir_upr*N)
mu.med<-xts(c(dat_csv$totale_attualmente_positivi,I_seir_med*N),order.by = c(days.before,days.ahead),frequency = 7)
counts<-mu.med
mu<-xts(x = as.matrix(cbind(counts,mu.lower,mu.upper)) , order.by = c(days.before,days.ahead))
p <- dygraph(mu,main=paste("Italy: Scenario 1 Credible (confidence ",100*0.90,"%)",sep = ""),ylab=" Infected",xlab="Day",height=400,width=800) %>%  dySeries(c("mu.lower", "counts", "mu.upper"),label="counts")
p<-p %>% dyLegend(show = "always", hideOnMouseOut = FALSE) %>%  dyShading(from = days.ahead[1], to = days.ahead[step.ahead], color = "#CCEBD6")%>% dyEvent(days.ahead[1], "Prediction", labelLoc = "bottom")
p

### confidence limits - scenario 2
I_seir_med<-tapply(seir2_sim$I,seir2_sim$time,median)
I_seir_lwr<-tapply(seir2_sim$I,seir2_sim$time,quantile,p=0.05)
I_seir_upr<-tapply(seir2_sim$I,seir2_sim$time,quantile,p=0.95)

mu.lower<-c(dat_csv$totale_attualmente_positivi,I_seir_lwr*N)
mu.upper<-c(dat_csv$totale_attualmente_positivi,I_seir_upr*N)
mu.med<-xts(c(dat_csv$totale_attualmente_positivi,I_seir_med*N),order.by = c(days.before,days.ahead),frequency = 7)
counts<-mu.med
mu<-xts(x = as.matrix(cbind(counts,mu.lower,mu.upper)) , order.by = c(days.before,days.ahead))
p <- dygraph(mu,main=paste("Italy: Scenario 2 Credible (confidence ",100*0.90,"%)",sep = ""),ylab=" Infected",xlab="Day",height=400,width=800) %>%  dySeries(c("mu.lower", "counts", "mu.upper"),label="counts")
p<-p %>% dyLegend(show = "always", hideOnMouseOut = FALSE) %>%  dyShading(from = days.ahead[1], to = days.ahead[step.ahead], color = "#CCEBD6")%>% dyEvent(days.ahead[1], "Prediction", labelLoc = "bottom")
p

### confidence limits - scenario 3
I_seir_med<-tapply(seir3_sim$I,seir3_sim$time,median)
I_seir_lwr<-tapply(seir3_sim$I,seir3_sim$time,quantile,p=0.05)
I_seir_upr<-tapply(seir3_sim$I,seir3_sim$time,quantile,p=0.95)

mu.lower<-c(dat_csv$totale_attualmente_positivi,I_seir_lwr*N)
mu.upper<-c(dat_csv$totale_attualmente_positivi,I_seir_upr*N)
mu.med<-xts(c(dat_csv$totale_attualmente_positivi,I_seir_med*N),order.by = c(days.before,days.ahead),frequency = 7)
counts<-mu.med
mu<-xts(x = as.matrix(cbind(counts,mu.lower,mu.upper)) , order.by = c(days.before,days.ahead))
p <- dygraph(mu,main=paste("Italy: Scenario 3 Credible (confidence ",100*0.90,"%)",sep = ""),ylab=" Infected",xlab="Day",height=400,width=800) %>%  dySeries(c("mu.lower", "counts", "mu.upper"),label="counts")
p<-p %>% dyLegend(show = "always", hideOnMouseOut = FALSE) %>%  dyShading(from = days.ahead[1], to = days.ahead[step.ahead], color = "#CCEBD6")%>% dyEvent(days.ahead[1], "Prediction", labelLoc = "bottom")
p
```

>>>>

---
title: "COVID19 - Forecast and predictions using a time dependent SEIR model"
author: "Paolo Girardi"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::github_document
always_allow_html: yes
---
```{r echo=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir= normalizePath('..'))
knitr::opts_chunk$set(error = FALSE)
```
<a rel="license" href="http://creativecommons.org/licenses/by-nc/4.0/"><img alt="Creative Commons License" style="border-width:0" src="https://i.creativecommons.org/l/by-nc/4.0/88x31.png" /></a><br />This work is licensed under a <a rel="license" href="http://creativecommons.org/licenses/by-nc/4.0/">Creative Commons Attribution-NonCommercial 4.0 International License</a>.

@@

NB:  the file draft_analysis_italy_html.Rmd performs the same analysis enabling Javascript Pictures.
## The COVID dataset

@@

```{r echo=TRUE}
checkpackage <- function(package) {
  if (!package %in% installed.packages()) install.packages(package)
}
checkpackage("dygraphs")
checkpackage("xts")
checkpackage("EpiDynamics")
checkpackage("webshot")
checkpackage("bsts")
checkpackage("ggplot2")
```
and load them.
```{r, warning=FALSE}
library(dygraphs)
library(xts)
library(EpiDynamics)
library(webshot)
library(bsts)
library(EpiDynamics)
library(ggplot2)
```

@@

We calculate several R0 values, each one based on a mobile window of 5 days, that can be sufficient to estimate a local trend, in order to assess if the R0 trend is decreasing (how is expected to be). In this way, the R0 for the first and the last two days of observation is impossibile to estimate.

@@

fp <- ggplot(data=df, aes(x=label, y=mean, ymin=lower, ymax=upper)) +
  geom_pointrange() +
  geom_hline(yintercept=1, lty=2) +  # add a dotted line at x=1 after flip
  xlab("Date") + ylab("R0 Mean (95% CI)") +
  theme_bw() 
print(fp)

@@

The R0 shows a decreasing trend in the last period. We use the estimated trend between R0 and time to calculate the future R0 value for the next 14 days. 
We predict beta (and R0) for the next 14 days by means of a linear regressione model, assuming a Log-normal distribution for the beta (the slope) and forcing its value to be greater than 0.
```{r R0 forecast}

@@

<img src="http://www.public.asu.edu/~hnesse/classes/seireqn.png"/>  
<img src="https://upload.wikimedia.org/wikipedia/commons/3/3d/SEIR.PNG"/>  
Respect to the previous formulation, the compartment of exposed people (E) was inserted between the susceptible and infected compartments.  
(https://en.wikipedia.org/wiki/Bayesian_structural_time_series)

We perfom simulation on the trend of beta coefficient by means of a Bayesian Structural Time Series using the library bsts of R.    
We estimate a BSTS model specifing a local linear trend ans 1000 simulation.

We made a forecast forward of 14 days dropping the first 100 simulation as burn-in.

```{r bsts model fit and prediction, echo=TRUE}
# Bayesian Structural Time Series
set.seed(123)
ss <- AddLocalLinearTrend(list(), log(beta_vec))
model1 <- bsts(log(beta_vec),
               state.specification = ss,
               niter = 1000)
par(mfrow = c(1,1))
plot(model1, "components", ylab="Log beta coefficient", xlab="days")
#previsioni
pred1 <- predict(model1, horizon = 16, burn = 100)
par(mfrow = c(1,1))
plot(pred1 , ylab="Log beta coefficient",main="Data and predictions")
# matrix of beta coefficients
coef<-(exp(pred1$distribution)[,3:16])
par(mfrow = c(1,1))
```
For each vectors of simulated beta coefficients we perform a SEIR model (one for each scenario).  
We save the results and plot a credible interval at 90% (5-95% percentile).

```{r  scenario plot , echo=TRUE}

seir1_sim<-seir2_sim<-seir3_sim<-NULL
for(s in 1:dim(coef)[1]){
  # average number of single connections of an infected person
  # less contacts, less probability of new infections
  # we keep constant the other parameters
  forecast<-14
  seir1<-seir2<-seir3<-NULL
  for(i in 1:forecast){
    parameters <- c(mu = mu0, beta = matrix(coef[s,i]), sigma = sigma0, gamma = 1/duration)
    f1<-10
    if( i==1) initials <- c(S = 0.95, E = (f1*I0/N), I = I0/N, R = R0/N)
    if( i>1) initials <- c(S = seir1_temp$results$S[2], E = seir1_temp$results$E[2], I =seir1_temp$results$I[2], R = seir1_temp$results$R[2])
    seir1_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
    seir1 <- rbind(seir1,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
    f2<-5
    if( i==1) initials <- c(S = 0.95, E = (f2*I0/N), I = I0/N, R = R0/N)
    if( i>1) initials <- c(S = seir2_temp$results$S[2], E = seir2_temp$results$E[2], I =seir2_temp$results$I[2], R = seir2_temp$results$R[2])
    seir2_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
    seir2 <- rbind(seir2,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
    f3<-3
    if( i==1) initials <- c(S = 0.95, E = (f3*I0/N), I = I0/N, R = R0/N)
    if( i>1) initials <- c(S = seir3_temp$results$S[2], E = seir3_temp$results$E[2], I =seir3_temp$results$I[2], R = seir3_temp$results$R[2])
    seir3_temp <- SEIR(pars = parameters, init = initials, time = 0:1)
    seir3 <- rbind(seir3,SEIR(pars = parameters, init = initials, time = 0:1)$results[2,])
  }
  seir1_sim<-rbind(seir1_sim,cbind(rep(s,forecast),seir1))
  seir2_sim<-rbind(seir2_sim,cbind(rep(s,forecast),seir2))
  seir3_sim<-rbind(seir3_sim,cbind(rep(s,forecast),seir3))
}
seir1_sim[,2]<-seir2_sim[,2]<-seir3_sim[,2]<-rep(1:forecast,dim(coef)[1])
colnames(seir1_sim)[1]<-colnames(seir2_sim)[1]<-colnames(seir3_sim)[1]<-"sim"

### confidence limits
I_seir_med<-tapply(seir1_sim$I,seir1_sim$time,median)
limits<-c(0.05,0.95)
confidence<-diff(limits)
I_seir_lwr<-tapply(seir1_sim$I,seir1_sim$time,quantile,p=limits[1])
I_seir_upr<-tapply(seir1_sim$I,seir1_sim$time,quantile,p=limits[2])

days.before<-date[1:days]
days.ahead<-date[(days+1):(days+forecast)]
step.ahead<-forecast+1
mu.lower<-c(dat_csv$totale_attualmente_positivi,I_seir_lwr*N)
mu.upper<-c(dat_csv$totale_attualmente_positivi,I_seir_upr*N)
mu.med<-xts(c(dat_csv$totale_attualmente_positivi,I_seir_med*N),order.by = c(days.before,days.ahead),frequency = 7)
counts<-mu.med
mu<-xts(x = as.matrix(cbind(counts,mu.lower,mu.upper)) , order.by = c(days.before,days.ahead))
p <- dygraph(mu,main=paste("Italy: Scenario 1 (Credible Interval ",100*confidence,"%)",sep = ""),ylab=" Infected",xlab="Day",height=400,width=800) %>%  dySeries(c("mu.lower", "counts", "mu.upper"),label="counts")
p<-p %>% dyLegend(show = "always", hideOnMouseOut = FALSE) %>%  dyShading(from = days.ahead[1], to = days.ahead[step.ahead], color = "#CCEBD6")%>% dyEvent(days.ahead[1], "Prediction", labelLoc = "bottom")
p

### confidence limits - scenario 2
I_seir_med<-tapply(seir2_sim$I,seir2_sim$time,median)
I_seir_lwr<-tapply(seir2_sim$I,seir2_sim$time,quantile,p=limits[1])
I_seir_upr<-tapply(seir2_sim$I,seir2_sim$time,quantile,p=limits[2])

mu.lower<-c(dat_csv$totale_attualmente_positivi,I_seir_lwr*N)
mu.upper<-c(dat_csv$totale_attualmente_positivi,I_seir_upr*N)
mu.med<-xts(c(dat_csv$totale_attualmente_positivi,I_seir_med*N),order.by = c(days.before,days.ahead),frequency = 7)
counts<-mu.med
mu<-xts(x = as.matrix(cbind(counts,mu.lower,mu.upper)) , order.by = c(days.before,days.ahead))
p <- dygraph(mu,main=paste("Italy: Scenario 2 (Credible Interval ",100*confidence,"%)",sep = ""),ylab=" Infected",xlab="Day",height=400,width=800) %>%  dySeries(c("mu.lower", "counts", "mu.upper"),label="counts")
p<-p %>% dyLegend(show = "always", hideOnMouseOut = FALSE) %>%  dyShading(from = days.ahead[1], to = days.ahead[step.ahead], color = "#CCEBD6")%>% dyEvent(days.ahead[1], "Prediction", labelLoc = "bottom")
p

### confidence limits - scenario 3
I_seir_med<-tapply(seir3_sim$I,seir3_sim$time,median)
I_seir_lwr<-tapply(seir3_sim$I,seir3_sim$time,quantile,p=limits[1])
I_seir_upr<-tapply(seir3_sim$I,seir3_sim$time,quantile,p=limits[2])

mu.lower<-c(dat_csv$totale_attualmente_positivi,I_seir_lwr*N)
mu.upper<-c(dat_csv$totale_attualmente_positivi,I_seir_upr*N)
mu.med<-xts(c(dat_csv$totale_attualmente_positivi,I_seir_med*N),order.by = c(days.before,days.ahead),frequency = 7)
counts<-mu.med
mu<-xts(x = as.matrix(cbind(counts,mu.lower,mu.upper)) , order.by = c(days.before,days.ahead))
p <- dygraph(mu,main=paste("Italy: Scenario 3 (Credible Interval ",100*confidence,"%)",sep = ""),ylab=" Infected",xlab="Day",height=400,width=800) %>%  dySeries(c("mu.lower", "counts", "mu.upper"),label="counts")
p<-p %>% dyLegend(show = "always", hideOnMouseOut = FALSE) %>%  dyShading(from = days.ahead[1], to = days.ahead[step.ahead], color = "#CCEBD6")%>% dyEvent(days.ahead[1], "Prediction", labelLoc = "bottom")
p
```
